---
title: "Simulating the start of an Ebola outbreak"
author: Thibaut Jombart
date: "`r format(Sys.time(), '%d %B, %Y')`"
categories: ["R"]
tags: ["R Markdown", "plot", "regression"]
output:
  html_document:
    toc: true
    toc_depth: 3
---

	
```{r options, include = FALSE, message = FALSE, warning = FALSE, error = FALSE}
library(knitr)
opts_chunk$set(collapse = TRUE)

install_if_missing <- function(x) {
    if (!require(x, character.only = TRUE)) {
        install.packages(x)
        require(x, character.only = TRUE)
    }
}

deps <- c("incidence", "epitrix", "distcrete", "projections")
lapply(deps, install_if_missing)

```



# Using data from past outbreak: the serial interval

We create a distribution of the serial interval (SI), i.e. the delays between
primary and secondary symptom onsets, in Ebola Virus Disease (EVD). We start by
converting previously estimates of the mean and standard deviation of the SI
(WHO Ebola Response Team (2014) NEJM 371:1481–1495) to the parameters of a Gamma
distribution:

```{r si-params}
library(epitrix)

mu <- 15.3 # mean in days days
sigma <- 9.3 # standard deviation in days
cv <- sigma/mu # coefficient of variation
cv
param <- gamma_mucv2shapescale(mu, cv) # convertion to Gamma parameters
param

```

The *shape* and *scale* are parameters of a Gamma distribution we can use to
generate delays. However, delays are typically reported per days, which implies
a discretisation (from continuous time to discrete numbers). We use the package
[*distcrete*](https://github.com/reconhub/distcrete) to achieve this
discretisation:

```{r si}

si <- distcrete::distcrete("gamma", interval = 1,
               shape = param$shape,
               scale = param$scale, w = 0)
si

plot(0:60, si$d(0:60), type = "h", xlab = "Days after onset",
     ylab = "P(secondary symptoms)", lend = 2, lwd = 4, col = "#666699",
     main = "Serial interval distribution")

```



# Simulation using a Poisson branching process

## Simulations from a single initial case

We use the simulations implemented in `projections` to simulate the start (first
60 days) of an Ebola outbreak (see
[here](http://www.repidemicsconsortium.org/projections/) for details of model
and implementation). Here we use the serial interval defined above, and
arbitrary possible values of $R$ from 1.1 to 3.5, each of which uses 1,000
simulations:

```{r projections}
library(projections)

R_values <- c(1.1, 1.5, 2, 2.5, 3, 3.5)

res <- lapply(R_values,
              function(R)
                project(incidence(0), R = R, si = si, n_days = 60, n_sim = 1000)
              )
names(res) <- R_values


```

There are different ways to look at these results, but one possible summary
would be to look at the total number of cases accummulating over the first 60
days, for each simulation:

```{r cumul-count}

count_cumul_cases <- function(x) {
  out <- apply(x, 2, cumsum)
  class(out) <- c("cumul_cases", "matrix")
  out
}

plot.cumul_cases <- function(x, ...) {

  coords <- boxplot(t(x), col = "#862d59", xlab = "Days from onset of index case",
                    ylab = "Cumulative number of cases", outline = FALSE, range = Inf,
                    ...)

  quantiles <- apply(x, 1, quantile, c(.95, .99))
  for (i in 1:nrow(quantiles)) {
    points(1:ncol(quantiles), quantiles[i, ], type = "l", col = "#862d59")
  }
}

cumul_cases <- lapply(res, count_cumul_cases)

```

In the following graph, we represent the cumulative number of cases from 1,000
simulations, up to 60 days after the onset of the first case, assuming various
values of the reproduction number. Outliers are not displayed on these barplots
but the maximum alues are indicated, as well as the 95% and 99% quantiles
(lines).

```{r options-wide-figures, echo = FALSE}
opts_chunk$set(fig.height = 5, fig.width = 10, out.width = "100%")
```

```{r final-plots}

par(mar = c(5.1, 4.1, 4.1, 0.1))

for (i in 1:length(cumul_cases)) {
  plot(cumul_cases[[i]], main = paste("Assumed R:", names(res)[i]))
}

```


**In short**: for moderate reproduction numbers, it is unlikely that a **single
index case** would lead to more than a **handful of cases** over the first 2
months. For larger $R_0$ (2.5 and above) we cannot exclude more substantial
epidemics (50 to a 200-300). Again, this relies **solely on a single case**. The
current situation is likely more concerning, as there are 2 confirmed cases (see
next section for sensitivity).



## Using a few cases as initial conditions

This is very much a 'what if' scenario, as we don't currently have daily
incidence. Still, we can replicate the simulations using more cases, distributed
randomly over a given time period. 

The function below will achieve these simulations, using `n_rep` simulated time
series as input (default 100), simulating `n_sim` random trajectories for each
time series, and pulling the results together.

```{r make_simul}

make_simul <- function(n_initial_cases = 1, time_period = 30, n_rep = 100, ...) {
  out <- vector(length = n_rep, mode = "list")
  for (i in seq_len(n_rep)) {
    fake_data <- c(0, # index case
                   sample(seq_len(time_period),
                          n_initial_cases - 1,
                          replace = TRUE)
                   )
    fake_epicurve <- incidence::incidence(fake_data)
    out[[i]] <- project(fake_epicurve, ...)
  }

  do.call(cbind, out)
}

```

In the following subsections, we give examples of 'what if' scenarios, for
different number of initial cases and time windows.


### Expected number of cases if 2 cases occured on the same week ($R = 2$)

```{r 2-cases-same-week}

x_2_7 <- make_simul(2, 7, R = 2, si = si, n_days = 60, n_sim = 100)
x_2_7_cumul <- count_cumul_cases(x_2_7)
plot(x_2_7_cumul)

```


### Expected number of cases if 2 cases occured on the same month ($R = 2$)

```{r 2-cases-same-month}

x_2_30 <- make_simul(2, 30, R = 2, si = si, n_days = 60, n_sim = 100)
x_2_30_cumul <- count_cumul_cases(x_2_30)
plot(x_2_30_cumul)

```


